Self-tunable approximated explicit MPC: Heat exchanger implementation and analysis

### 2023-06-19 paper/ATE/main.tex
ATE update1

\journal{Energy}

@@

			The tunable approximated explicit model predictive control (MPC) comes with the benefits of real-time tunability without the necessity of solving the optimization problem online. This paper provides a novel self-tunable control policy that does not require any further tuning or interventions of the control engineer during operation. Based on the current operating conditions, the autonomous tuning parameter scales the control input using linear interpolation between the boundary optimal control actions. The adjustment of the tuning parameter depends on the current reference value, which makes this strategy suitable for reference tracking problems. Furthermore, a novel technique for scaling the tuning parameter is proposed. It provides to exploit different ranges of the tuning parameter assigned to specified operating conditions. The self-tunable explicit MPC was implemented on a laboratory heat exchanger with nonlinear and asymmetric behavior. The asymmetric behavior of the plant was compensated by tuning the controller's aggressiveness, as the negative or positive sign of reference change was considered in the tuning process. The designed self-tunable controller improved control performance by decreasing integral squared error, maximal overshoots/undershoots and settling time compared to the conventional control strategy based on a single controller. 
			\end{abstract}
		%
		%%%Graphical abstract
		%\begin{graphicalabstract}
			%%\includegraphics{grabs}
			%\end{graphicalabstract}
		%
		%%Research highlights
		\begin{highlights}
			% 3-5 highlights, every one should not exceed 85 characters including spaces
			\item Self-tunable control technique does not need any intervention during control
			\item Tuning based on operating conditions compensates for the nonlinear plant behavior
			\item Implementation of the proposed tuning method on a laboratory heat exchanger 
			\item Control performance improved compared to utilizing only one explicit MPC
			\end{highlights}

@@

	\begin{figure}
		\begin{center}
			\includegraphics[width=\textwidth]{images/tuning}
			\caption[Self-tuning scheme]{Scheme of the self-tuning control evaluation.}
			\label{fig:tuning}
		\end{center}
	\end{figure}

@@

The plant on which the control was implemented and analyzed is a laboratory-scaled counter-current liquid-to-liquid plate heat exchanger Armfield Process Plant Trainer PCT23~\cite{pct23}, see Figure~\ref{fig:HE}. The heat exchanger is 90\,mm wide, 103\,mm long, and 160\,mm high. The heat exchange is performed between the cold feed (water) and the hot medium (water). The cold feed as well as the heating medium are transported to the heat exchanger by two peristaltic pumps with flexible tubing from silicon rubber. The flow rate of the feed is constant, while the aim of control is to track the reference value of the feed temperature. Therefore, the controlled variable is the feed temperature $T$ at the outlet of the heat exchanger. The temperature of the heated feed in the outlet stream was measured by the type K thermocouple. The associated manipulated variable is the voltage $U$ corresponding to the power of the pump feeding the heat exchanger by the hot medium. The voltage is within the range of [$0-5$]\,V normalized into the relative values in percentage. For further technical specifications of the laboratory heat exchanger, the reader is referred to~\cite{pct23}. As heat exchange is a nonlinear and asymmetric process~\cite{Liptak}, this heat exchanger represents a suitable candidate for the presented controller tuning strategy. The corresponding illustrative scheme of the implemented closed-loop control setup is in Figure~\ref{fig:CL}, where the ``Self-tuning'' block substitutes the more detailed scheme in Figure~\ref{fig:tuning}.

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\textwidth]{images/cl}
		\caption[Scheme of the implemented closed-loop control setup.]{Scheme of the implemented closed-loop control setup.}
		\label{fig:CL}
	\end{center}
\end{figure}

@@

These main experimental observations established the base for the strategy of self-tuning the penalty matrix $Q_\mathrm{y}$. The strategy follows the ideas summarized in Section~\ref{sec:methodology}. Therefore, utilizing the boundary controller with the penalty matrix $Q_\mathrm{y, L}$ is preferred when the reference changes upwards. On the contrary, the controller associated with $Q_\mathrm{y, U}$ is preferred for negative reference step changes. The splitting value of the tuning parameter was chosen in the middle of the interval, i.e., $\rho_{\mathrm{s}} = 0.5$. The remaining parameter that needed to be set was the maximal admissible size of the reference step change $\Delta_{\max}$, which was determined to $15\,^{\circ}\mathrm{C}$ as the investigated range of controlled temperature was [$35-50$]$^{\circ}\mathrm{C}$.

@@

	\subsection*{Abbreviations}
	%	\begin{center}
		\begin{tabular}{ l l }
			ISE  & integral squared error \\
			LTI  & linear time-invariant (system) \\
			MIMO & multiple-inputs and multiple-outputs (system) \\
			MPC  & model predictive control \\
			mp-QP& multi-parametric quadratic programming (problem) \\
			PID  & proportional–integral–derivative (controller) \\
			PWA  & piece-wise affine (function) \\
			PWC  & piece-wise constant (function) \\
			QP   & quadratic programming (problem) \\
			SISO & single-input and single-output (system) \\
		\end{tabular}
		%	\end{center}

>>>>

\journal{Applied Thermal Engineering}

@@

			The tunable approximated explicit model predictive control (MPC) comes with the benefits of real-time tunability without the necessity of solving the optimization problem online. This paper provides a novel self-tunable control policy that does not require any further tuning or interventions of the control engineer during operation. Based on the current operating conditions, the autonomous tuning parameter scales the control input using linear interpolation between the boundary optimal control actions. The adjustment of the tuning parameter depends on the current reference value, which makes this strategy suitable for reference tracking problems. Furthermore, a novel technique for scaling the tuning parameter is proposed. It provides to exploit different ranges of the tuning parameter assigned to specified operating conditions. The self-tunable explicit MPC was implemented on a laboratory heat exchanger with nonlinear and asymmetric behavior. The asymmetric behavior of the plant was compensated by tuning the controller's aggressiveness, as the negative or positive sign of reference change was considered in the tuning process. The designed self-tunable controller improved control performance by decreasing integral squared error, maximal overshoots/undershoots and settling time compared to the conventional control strategy based on a single controller. 
			\end{abstract}
		%
		%%%Graphical abstract
		%\begin{graphicalabstract}
			%%\includegraphics{grabs}
			%\end{graphicalabstract}
		%
		%%Research highlights
		\begin{highlights}
			% 3-5 highlights, every one should not exceed 85 characters including spaces
			\item Self-tunable control technique does not need any intervention during control
			\item Tuning based on operating conditions compensates for the nonlinear plant behavior
			\item Implementation of the proposed tuning method on a laboratory heat exchanger 
			\item Control performance improved compared to utilizing explicit MPC with constant setup
			\end{highlights}

@@

	\begin{figure}
		\begin{center}
			\includegraphics[width=\textwidth]{images/tuning2}
			\caption[Self-tuning scheme]{Scheme of the self-tuning control evaluation.}
			\label{fig:tuning}
		\end{center}
	\end{figure}

@@

The plant on which the control was implemented and analyzed is a laboratory-scaled counter-current liquid-to-liquid plate heat exchanger Armfield Process Plant Trainer PCT23~\cite{pct23}, see Figure~\ref{fig:HE}. The heat exchanger is 90\,mm wide, 103\,mm long, and 160\,mm high. The heat exchange is performed between the cold feed (water) and the hot medium (water). The cold feed as well as the heating medium are transported to the heat exchanger by two peristaltic pumps with flexible tubing from silicon rubber. The flow rate of the feed is constant, while the aim of control is to track the reference value of the feed temperature. Therefore, the controlled variable is the feed temperature $T$ at the outlet of the heat exchanger. The temperature of the heated feed in the outlet stream was measured by the type K thermocouple. The associated manipulated variable is the voltage $U$ corresponding to the power of the pump feeding the heat exchanger by the hot medium. The voltage is within the range of [$0-5$]\,V normalized into the relative values in percentage. The maximal voltage 5\,V or 100\% corresponds to flow rate 11.5\,$\mathrm{ml}\,\mathrm{s}^{-1}$. For further technical specifications of the laboratory heat exchanger, the reader is referred to~\cite{pct23}. As heat exchange is a nonlinear and asymmetric process~\cite{Liptak}, this heat exchanger represents a suitable candidate for the presented controller tuning strategy. The corresponding illustrative scheme of the implemented closed-loop control setup is in Figure~\ref{fig:CL}, where the ``Self-tuning'' block substitutes the more detailed scheme in Figure~\ref{fig:tuning}.

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\textwidth]{images/cl2}
		\caption[Scheme of the implemented closed-loop control setup.]{Scheme of the implemented closed-loop control setup, where ``eMPC'' denotes explicit MPC.}
		\label{fig:CL}
	\end{center}
\end{figure}

@@

These main experimental observations established the base for the strategy of self-tuning the penalty matrix $Q_\mathrm{y}$. The strategy follows the ideas summarized in Section~\ref{sec:methodology}. Therefore, utilizing the boundary controller with the penalty matrix $Q_\mathrm{y, L}$ is preferred when the reference changes upwards. On the contrary, the controller associated with $Q_\mathrm{y, U}$ is preferred for negative reference step changes. The splitting value of the tuning parameter was chosen simply in the middle of the interval, i.e., $\rho_{\mathrm{s}} = 0.5$. The remaining parameter that needed to be set was the maximal admissible size of the reference step change $\Delta_{\max}$, which was determined to $15\,^{\circ}\mathrm{C}$ as the investigated range of controlled temperature was [$35-50$]$^{\circ}\mathrm{C}$.

@@

	\subsection*{Abbreviations}
	%	\begin{center}
		\begin{tabular}{ l l }
			eMPC & explicit model predictive control \\
			ISE  & integral squared error \\
			LTI  & linear time-invariant (system) \\
			MIMO & multiple-inputs and multiple-outputs (system) \\
			MPC  & model predictive control \\
			mp-QP& multi-parametric quadratic programming (problem) \\
			PID  & proportional–integral–derivative (controller) \\
			PWA  & piece-wise affine (function) \\
			PWC  & piece-wise constant (function) \\
			QP   & quadratic programming (problem) \\
			SISO & single-input and single-output (system) \\
		\end{tabular}
		%	\end{center}
